Identifying optimal baselines for stable isotope studies of stream food webs

### 2022-05-05 docs/ms.Rmd
ms updates

---
title: "Identifying optimal baselines for stable isotope studies of stream food webs"
author: 
  - "Nathan Barrus^1^, Bryan M. Maitland^1,2^, Frank J, Rahel^1^"
  - "^1^University of Wyoming" 
  - "^2^UW-Madison and WDNR"
date: "`r format(Sys.time(), '%Y-%B-%d')`"
output: 
  bookdown::html_document2:
    number_sections: false
fontsize: 11pt
geometry: margin = 1in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage[left]{lineno}
- \linenumbers
- \usepackage{dcolumn}
- \usepackage{caption}
- \usepackage{float}
- \usepackage{afterpage}
- \usepackage{siunitx}
- \usepackage{amsmath}
bibliography: baseline-isotopes.bib
csl: freshwater-biology.csl
---

@@

Stable isotope analysis (SIA) is used in aquatic ecosystems to reconstruct food webs and deduce trophic interactions. Because land use strongly influence δ15N values in consumers, suitable isotopic baselines must be established for meaningful cross-site comparisons of food webs. 
Our objective was to identify an invertebrate taxon or Functional Feeding Group (FFG) useful as a baseline for δ15N values in riverine food webs. The ideal baseline should meet four criteria: 1) be easy to collect with wide geographic distribution, 2) have low within-site δ15N variation among individuals, 3) have δ15N values that are correlated with land use and background δ15N values, and 4) produce no systemic change in calculated consumer trophic positions with % land use (i.e. be independent of system-specific δ15N variability).  We investigated these criteria in invertebrate communities in 16 sites along a gradient from mountain to prairie streams which also represent a gradient in land use.

Two invertebrate families, Simuliidae and Ephemeridae, and the FFG of Filterers fulfilled all four criteria, similar to ideal indicator taxa identified in Danish lowland streams. Interestingly, even after correcting for local variation in baseline δ15N values, the trophic position of some fish species still varied along the longitudinal stream gradient. This difference is most likely explained by actual differences in trophic positions of these species due to dietary changes across the stream gradient.

@@

Problem: δ15N values vary in freshwater systems, e.g., reflecting variations in land use, so suitable baseline indicators must be specified to permit cross-site comparisons, and few methodical SOPs have been developed.

@@

Stable isotope analysis (SIA) is a commonly-used tool for making inferences about the trophic structure and flow of materials in food webs [@Perkins2014; @VanderZanden1997]. Stable isotope ratios—typically of carbon (13C/12C, δ13C) or nitrogen (15N/14N, δ15N)—provide information on the materials assimilated by animals because of predictable changes in isotopic composition that occur in body tissues during elemental incorporation of diet items [@VanderZanden2001]. With each trophic transfer, carbon isotope ratios change little which makes them useful for tracking energy pathways from basal resources and identifying the ultimate carbon source supporting consumers. In contrast, nitrogen isotope ratios change markedly and predictably with each trophic transfer, allowing for inference into species’ trophic positions in communities, as well quantification of whole-community food chain length (Post, 2002; Mccutchan et al., 2016; Pinzone et al., 2017).

However, large spatial variability in the nitrogen isotope ratios—that is, δ15N values—of aquatic organisms can make the comparison of food webs among sites or along environmental  gradients problematic (Anderson & Cabana, 2005; Barnes et al., 2008).  Such variability arises from natural processes such as denitrification and nitrogen fixation, but also and importantly from land use changes (Chappuis et al., 2017).  Human-dominated landscapes increase the number of denitrified components of nitrogen in the system (i.e. ammonium and POM) through bacterial decomposition of anthropogenic sources of nitrogen (i.e. agriculture, sewage treatment, livestock manure). Because bacteria preferentially use 14N over 15N when breaking down these nitrogen components the remaining nitrogen from these components that enter the system have elevated δ15N signatures (Di Lascio et al., 2013). The δ15N values from anthropogenic sources of nitrogen have been shown to propagate up the food web. For example, in the Saint Lawrence lowlands in Quebec, Canada, primary consumer δ15N values increased with the intensity of regional nitrogen loading from synthetic fertilizers and pig, chicken and cattle manure (Anderson & Cabana, 2006).  In the northeastern Unites States, watersheds dominated by natural forests had lower δ15N values than watersheds dominated by urban or agricultural land (Mayer et al., 2002).  And in streams in northeastern Spain, δ15N values of all basal resources were highest in human-impacted mainstem locations (Pastor et al., 2013).  In addition, land-use practices have been shown to elevate δ15N values in lacustrine systems in Rhode Island (Lake et al., 2001) and boreal streams in south-central Sweden (Bergfur et al., 2009).  Thus, comparing food webs from sites dominated by different land uses can cause erroneous inferences into food web structure and function.

More accurate inferences can be achieved by applying a correction to consumer isotope ratio data that accounts for local variability in δ15N among sample sites.  Trophic Position (TP) in particular is sensitive to variability in δ15N and requires correction for better inferences (Kristensen et al., 2016). Such a correction can be achieved by modeling the δ15N variability of all primary consumers in a system (Vander-Zanden & Rasmussen, 1999).  But the assessment of TP for higher level consumers can be confounded by inter-site difference in the δ15N of basal resources (Hussey et al., 2014). A more promising approach is to develop a correction using an organism as a baseline proxy that reflects background δ15N variation among sites (Cabana and Rasmussen 1996, Post 2002, Kristensen et al. 2016, Quezada-Romegialli et al. 2018) and that has a TP close to the higher level consumers (Hussey et al., 2014).  Although this baseline method is beginning to be applied in food web studies, it has limitations which require careful consideration before a proxy baseline is selected. Primary producers have short life cycles and high tissue turnover rates which prohibits useful interpretation of higher trophic levels due to differing temporal resolutions. So researchers have tended to collect all potential basal primary resources at a site (i.e., suspended and benthic organic matter, biofilms, filamentous algae, macrophytes, and riparian vegetation) in replicate multiple times across a given time interval (e.g., three times over growing period) and use the average δ15N values of the primary producers as the baseline proxy value (Vinagre et al., 2008; Govender, Smit & Perissinotto, 2011). But collecting primary producers in streams is arduous and expensive because it requires many samples over many dates to adequately characterize baseline variation.  Therefore, long-lived primary consumers, particularly bivalves, with tissue turnover rates closer to higher level consumers are more appropriate than primary producers (Vander-Zanden & Rasmussen, 2001; Jardine et al., 2014). Bivalves also exhibit a uniform, specialized filter feeding strategy which can reduce additional variation in δ15N attributable to omnivorous feeding habits. Unfortunately, bivalves and similar long-lived primary consumers are sparsely distributed in many temperate streams, rendering them impractical for such systems.

Suitable primary consumer taxa to use as baseline indicators in temperate streams have only been identified for Danish lowlands and for an eastern Canadian watershed (Anderson & Cabana, 2007; Kristensen et al., 2016). But few detailed comparative tests exist, and a consensus needs to be reached for baseline indicators across regions. Therefore, our objective was to compare macroinvertebrate taxa and groups of taxa (i.e., functional feeding group; FFGs) as baseline indicators that reflect δ15N values at the base of food web in streams that span the Rocky Mountain—Great Plaines ecotone in western North America. These streams vary in land-use characteristics moving from upstream headwaters to downstream reaches, making them opportune systems to identify a widely useable and reliable baseline indicator. We used four criteria to identify an optimal invertebrate baseline (Kristensen et al., 2016). The baseline taxon must: (1) have a wide geographic distribution, (2) have low with-in site variation of δ15N values, (3) have δ15N values that are correlated with background spatial variability in δ15N, and (4) give trophic position estimates of organisms higher in the food web that are free of environment influences of δ15N. As urbanization and agricultural land-use practices can influence δ15N values in aquatic ecosystems (Mayer et al., 2002; Anderson & Cabana, 2005), the identification of regionally-derived baseline indicator organisms should aid in quantifying fish trophic position in stable isotope studies in lotic systems.

@@

We sampled sixteen 2nd-5th order stream sites distributed along the Rocky Mountain—Great Plains ecotone. Sites were selected to represent an environmental gradient based on elevation and land use that resulted in a wide variation of δ15N values at the base of local food webs. Sites spanned the longitudinal (i.e., upstream-downstream) gradient of the three primary tributaries to the North Platte River in Wyoming, USA (Figure 1). The Sweetwater, Medicine Bow and Laramie rivers are snowmelt dominated with peak flows occurring in the spring and base flows occurring in mid-summer. Their watersheds transition from forested land at high elevations to agriculturally dominated land with some urbanization at low elevations.

@@

Instead of using percent natural land cover as a variable to compare δ15N values among potential baselines (Kristensen et al., 2016), we used a proxy variable for a site’s location along the longitudinal stream gradient developed by Maitland (2020). The proxy variable combines seven metrics of environmental conditions at a study site into an index of longitudinal stream position using the Axis 1 site score from a principal component analysis (PCA). Input variables to the PCA were elevation (m a.s.l.), stream slope (km/km), distance to North Platte River (km), Strahler stream order, mean August water temperature (℃), upstream drainage area (km2), and stream width (m). This longitudinal gradient index (PC1) explained 78% of site-level variation among the 16 study sites and described changes in environmental conditions from upstream, high-elevation, cold-water streams to downstream, low-elevation, warm-water streams. Upstream sites were more likely to be surrounded by natural land cover (e.g. forest, grassland, shrubland, and wetland land types; NLCD Database 2011), as evidenced by a strong negative correlation between PC1 and percent natural landcover at a site (Pearson's correlation, r = -0.69, t15 = -3.66, P < 0.001; Maitland, 2020). Maitland (2020) also found that PC1 (and thus the degree of natural landcover) was positively correlated with δ15N values of basal resources and primary consumers (Linear Regression - Basal Resources: F1,373 = 121.4, P < 0.001; Primary Consumers: F¬¬1,14 = 4.7, P = 0.050), consistent with previous research (Cabana and Rasmussen 1996, Lake et al. 2001, Kristensen et al. 2016). Thus, we used PC1 as a measure of local variation in δ15N values for all future analyses. 
Sample collection and preparation

Samples were collected during summer 2016. Basal resources (i.e., primary uptake compartments) were sampled once in June, July, and August to account for fine-scale temporal variation in δ15N (Jardine et al., 2014).  Seston was collected by filtering three replicates of up to 10 L of water onto pre-combusted (550 °C, 4 hours) filters (Whatman GF/F, 47 mm O) in the field using a modified portable drill pump (Kelso & Baker, 2016). Biofilm was collected by scraping five cobble-sized rocks within the reach.  Macroinvertebrates were collected during the same sampling events during the summer of 2016. A D-framed kick net was used in a variety of habitats (i.e. riffle, pools, macrophytes, and river margins) to collect a representative sample of the macroinvertebrate assemblage.  Macroinvertebrates were kept in filtered stream water for 24 hours to allow for gut clearance, then frozen. Muscle tissue samples were obtained from fish collected during the August sampling event using a backpack electrofishing unit (Smith-Root, Vancouver, WA).  Brown Trout Salmo trutta (Linnaeus, 1758), Creek Chub Semotilus atromaculatus (Person, Year), White Sucker Catostomus commersonii (Lacépède, 1803), were used because these species were found at greater than 50% of the sites.  To minimize the number of fish euthanized, a muscle plug using a 5-mm biopsy punch was taken from fish greater than 300 mm in length. The remaining fish used for SIA were euthanized with a lethal dose of MS-222, placed on ice and then frozen upon returning to the lab. Those fish not selected for SIA were returned to the stream.

@@

When possible, we used individual macroinvertebrates for SIA but for small taxa we pooled individuals to ensure enough material for SIA. Muscle fillets were removed from the cranial dorsal portion of each fish. Samples were oven died (60 C, 48 hours), ground into a homogenous powder, and weighed to the nearest 0.001 mg. Animal samples (~1.0 mg) and basal resources (~2.0 mg) were weighed into 8 x 5-mm tin capsules and analyzed for 15N, 13C, %C, %N and C:N ratio at the University of Wyoming Stable Isotope Facility using a Delta Plus XP Continuous Flow Stable Isotope Ratio Mass Spectrometer (Thermo Finnigan, Bremen, Germany) coupled to a Costech Analytical 4010 elemental analyzer. Carbon (13C:12C) and nitrogen (15N:14N) isotope ratios (R) were estimated relative to their respective International Atomic Energy Agency standards (Pee Dee Belemnite limestone and atmospheric nitrogen). Working internal standards were run as controls throughout analysis to ensure accurate measurements. Isotopic values are reported in per mill delta notation \@ref(eq:ratio):

@@

where $\delta^{n}$ is either $N^{15}$ or $C^{13}$, and R is the ratio of the heavy to lighter isotope. Analytical error (i.e., 1 SD of lab standard) of sample runs was estimated at 0.07‰ ($\delta^{13}C$) and 0.18‰ ($\delta^{15}N$) for the invertebrates, and 0.04‰ ($\delta^{13}C$) and 0.10‰ ($\delta^{15}N$) for the fish.

@@

Suitability of various macroinvertebrate taxa as baselines were based on the four criteria proposed by Kristensen et al. (2016).  The baseline taxon must: (1) have a wide geographic distribution, (2) have low with-in site variation of δ15N values, (3) have δ15N values correlated with environmental influences of δ15N, (4) give trophic positions estimates of other consumers free of environment influences of δ15N.  Although all analyses were performed for all macroinvertebrate taxa and FFGs, selection occurred progressively from each criterion. For example, if a macroinvertebrate taxon met the first criterion it would stay in the list of potential baselines and be evaluated for the second criterion, but if it did not meet the first criterion then it would be removed from the list.

@@

We calculated the coefficient of variation (CV) of δ15N values for all macroinvertebrate taxa and FFGs. A low coefficient of variation for δ15N indicates that a baseline candidate has a non-omnivorous diet. To calculate CV, we divided the standard deviation of δ15N signatures by the mean δ15N signature of each baseline candidate at each site.  We then found the mean and 95% confidence intervals for each baseline candidate across sites.

@@

For all macroinvertebrate taxa and FFGs, we used least squares linear regression to assess relationships between the baseline’s δ15N position at sites along the longitudinal stream gradient.  We wished to compare our baselines’ relationships to the longitudinal gradient with those of the basal resources. So, we did separate least squares linear regressions for two of the basal resources at our sites and compared R2 values between the baselines and the resources at our sites.  The R2 value of our basal resources represent the minimum level of variation that our alternate baselines should explain in order to be used as alternatives to the traditional method (e.g. all basal resources across a season).   Baselines with significant (P ≤ 0.05) relationships with the longitudinal gradient and that had R2 values equal to or greater than the basal resources were considered suitable baselines.

@@

where TPConsumer is the trophic position of each individual fish at a site, δ15NConsumer is the δ15N signature of each individual fish, δ15NBaseline is the mean δ15N signature of the baseline at each site, ∆15N is the discrimination factor for each trophic transfer, and TLBaseline is the expected trophic level of the baseline. Discrimination factor were estimated using the methods of Caut et al. (2009).

@@

We calculated the coefficient of variation (CV) for δ15N at each site for the 12 sufficiently distributed baselines (8 taxa, 4 FFGs).  We found significant differences in the CVs for the taxa (ANOVA test; F11,26 = 2.96, P = 0.002; Figure 2). Simuliidae had the smallest CV across sites and Elmidae-adult had the largest (Figure 2).  Additionally, Simuliidae were 0.19 units lower than Elmidae-adult, 0.17 units lower than Chironimidae, and, although marginally significant, 0.15 units lower than Elmidae-larvae (Tukey HSD tests; P = 0.011 for Elmidae-adult; P = 0.032 for Chironomidae; P =  0.071 for Elmidae-adult; Figure 2).  Simuliidae were not statistically different from Dytiscidae, Baetidae, Heptaganeidae and Hydropyschidae. We found marginally significant differences among the FFG baselines (ANOVA test; F3,57 = 2.49, P = 0.070; Figure 2). Filterers were 0.18 units lower than Grazers (Tukey HSD tests; P = 0.047; Figure 2).  Filterers were not statistically different from Predators or Collectors.

### Criteria 3: Correlation with Environmental δ15N

The two basal resources, biofilm and seston, had δ15N signatures that were significantly (P ≤ 0.05) correlated with the longitudinal stream gradient (PC1) (Figure 3; Table 1). All eight of the well distributed taxa and all FFG’s had δ15N signatures that were significantly correlated with PC1 (Figure 3). The amount of variation in δ15N signatures captured by the longitudinal gradient was similar among the two basal resources, eight sufficiently distributed taxa and the four FFGs.

### Criteria 4: Trophic Position Estimates Independent of Environmental δ15N

@@

Similar to FFG’s, correction of the TP using different taxa significantly reduced the effect of the longitudinal gradient on TP estimates (Figure 5; Figure 7; Supplemental Table 1).  The slope of Brown TP estimates and PC1 was lower for all taxa than when using uncorrected estimates (Figure 7). After correction, Brown Trout TP estimates were positively correlated with the longitudinal gradient when using Hydropyschidae, Baetidae, Simulidae, Dytiscidae, Elmidae-larvae, and Elmidae-adult.  Brown Trout TP estimates were not correlated with the longitudinal gradient when using Heptaganeidae and Chironomidae as baselines. Brown Trout trophic position estimates differed by baseline taxa (ANOVA test: F7,2787 = 68.76; P < 0.001).   Brown Trout TP estimates given by Elmidae-adult and Elmidae-larvae were higher than the other taxa (Figure 6; Supplemental Table 2). Hydropyschidae and Heptaganeidae gave lower Brown Trout TP estimates than the other except Chironomidae and Baetidae for which they did not differ (Figure 6; Supplemental Table 2).  Dytiscidae and Simulidae gave intermediate Brown Trout TP estimates (Figure 6; Supplemental Table 2).

@@

Figure 2. Each baseline’s mean Coefficient of Variation (CV) of δ15N across sites with 90% confidence intervals. A) Taxon baselines B) FFG baselines.  All baselines met Criteria 1 of being found in at least 75% of study sites.

Figure 3. Relationships between the δ15N values of A) basal resources, B) FFGs, and C) taxa in relation to position along the longitudinal gradient represented by PC1.  Each point is the mean δ15N signature of the baseline candidate or basal resource at that site. Upstream sites are to the left. The black line is the fitted regression line and the shaded grey represents the 95% confidence interval.  See Supp. Table 1 for full statistical results.

>>>>

---
title: "Identifying optimal baselines for stable isotope studies of stream food webs"
author: 
  - "Nathan Barrus^1^, Bryan M. Maitland^1,2^, Frank J, Rahel^1^"
  - "^1^University of Wyoming" 
  - "^2^UW-Madison and WDNR"
date: "`r format(Sys.time(), '%Y-%B-%d')`"
output: 
  # word_document: default
  # pdf_document:
  #   fig_caption: yes
  #   keep_tex: yes
  #   latex_engine: pdflatex
  bookdown::html_document2:
    number_sections: false
fontsize: 11pt
geometry: margin = 1in
header-includes:
- \usepackage{setspace}\doublespacing
- \usepackage[left]{lineno}
- \linenumbers
- \usepackage{url}
- \usepackage{dcolumn}
- \usepackage{caption}
- \usepackage{float}
- \usepackage{afterpage}
- \usepackage{siunitx}
- \usepackage{amsmath}
# bibliography: baseline-isotopes.bib
# csl: freshwater-biology.csl
---

@@

Stable isotope analysis (SIA) is used in aquatic ecosystems to reconstruct food webs and deduce trophic interactions. Because land use strongly influence d15N values in consumers, suitable isotopic baselines must be established for meaningful cross-site comparisons of food webs. 
Our objective was to identify an invertebrate taxon or Functional Feeding Group (FFG) useful as a baseline for d15N values in riverine food webs. The ideal baseline should meet four criteria: 1) be easy to collect with wide geographic distribution, 2) have low within-site d15N variation among individuals, 3) have d15N values that are correlated with land use and background d15N values, and 4) produce no systemic change in calculated consumer trophic positions with % land use (i.e. be independent of system-specific d15N variability).  We investigated these criteria in invertebrate communities in 16 sites along a gradient from mountain to prairie streams which also represent a gradient in land use.

Two invertebrate families, Simuliidae and Ephemeridae, and the FFG of Filterers fulfilled all four criteria, similar to ideal indicator taxa identified in Danish lowland streams. Interestingly, even after correcting for local variation in baseline d15N values, the trophic position of some fish species still varied along the longitudinal stream gradient. This difference is most likely explained by actual differences in trophic positions of these species due to dietary changes across the stream gradient.

@@

Problem: d15N values vary in freshwater systems, e.g., reflecting variations in land use, so suitable baseline indicators must be specified to permit cross-site comparisons, and few methodical SOPs have been developed.

@@

Stable isotope analysis (SIA) is a commonly-used tool for making inferences about the trophic structure and flow of materials in food webs. Stable isotope ratios—typically of carbon (13C/12C, d13C) or nitrogen (15N/14N, d15N)—provide information on the materials assimilated by animals because of predictable changes in isotopic composition that occur in body tissues during elemental incorporation of diet items . With each trophic transfer, carbon isotope ratios change little which makes them useful for tracking energy pathways from basal resources and identifying the ultimate carbon source supporting consumers. In contrast, nitrogen isotope ratios change markedly and predictably with each trophic transfer, allowing for inference into species’ trophic positions in communities, as well quantification of whole-community food chain length (Post, 2002; Mccutchan et al., 2016; Pinzone et al., 2017).

However, large spatial variability in the nitrogen isotope ratios—that is, d15N values—of aquatic organisms can make the comparison of food webs among sites or along environmental  gradients problematic (Anderson & Cabana, 2005; Barnes et al., 2008).  Such variability arises from natural processes such as denitrification and nitrogen fixation, but also and importantly from land use changes (Chappuis et al., 2017).  Human-dominated landscapes increase the number of denitrified components of nitrogen in the system (i.e. ammonium and POM) through bacterial decomposition of anthropogenic sources of nitrogen (i.e. agriculture, sewage treatment, livestock manure). Because bacteria preferentially use 14N over 15N when breaking down these nitrogen components the remaining nitrogen from these components that enter the system have elevated d15N signatures (Di Lascio et al., 2013). The d15N values from anthropogenic sources of nitrogen have been shown to propagate up the food web. For example, in the Saint Lawrence lowlands in Quebec, Canada, primary consumer d15N values increased with the intensity of regional nitrogen loading from synthetic fertilizers and pig, chicken and cattle manure (Anderson & Cabana, 2006).  In the northeastern Unites States, watersheds dominated by natural forests had lower d15N values than watersheds dominated by urban or agricultural land (Mayer et al., 2002).  And in streams in northeastern Spain, d15N values of all basal resources were highest in human-impacted mainstem locations (Pastor et al., 2013).  In addition, land-use practices have been shown to elevate d15N values in lacustrine systems in Rhode Island (Lake et al., 2001) and boreal streams in south-central Sweden (Bergfur et al., 2009).  Thus, comparing food webs from sites dominated by different land uses can cause erroneous inferences into food web structure and function.

More accurate inferences can be achieved by applying a correction to consumer isotope ratio data that accounts for local variability in d15N among sample sites.  Trophic Position (TP) in particular is sensitive to variability in d15N and requires correction for better inferences (Kristensen et al., 2016). Such a correction can be achieved by modeling the d15N variability of all primary consumers in a system (Vander-Zanden & Rasmussen, 1999).  But the assessment of TP for higher level consumers can be confounded by inter-site difference in the d15N of basal resources (Hussey et al., 2014). A more promising approach is to develop a correction using an organism as a baseline proxy that reflects background d15N variation among sites (Cabana and Rasmussen 1996, Post 2002, Kristensen et al. 2016, Quezada-Romegialli et al. 2018) and that has a TP close to the higher level consumers (Hussey et al., 2014).  Although this baseline method is beginning to be applied in food web studies, it has limitations which require careful consideration before a proxy baseline is selected. Primary producers have short life cycles and high tissue turnover rates which prohibits useful interpretation of higher trophic levels due to differing temporal resolutions. So researchers have tended to collect all potential basal primary resources at a site (i.e., suspended and benthic organic matter, biofilms, filamentous algae, macrophytes, and riparian vegetation) in replicate multiple times across a given time interval (e.g., three times over growing period) and use the average d15N values of the primary producers as the baseline proxy value (Vinagre et al., 2008; Govender, Smit & Perissinotto, 2011). But collecting primary producers in streams is arduous and expensive because it requires many samples over many dates to adequately characterize baseline variation.  Therefore, long-lived primary consumers, particularly bivalves, with tissue turnover rates closer to higher level consumers are more appropriate than primary producers (Vander-Zanden & Rasmussen, 2001; Jardine et al., 2014). Bivalves also exhibit a uniform, specialized filter feeding strategy which can reduce additional variation in d15N attributable to omnivorous feeding habits. Unfortunately, bivalves and similar long-lived primary consumers are sparsely distributed in many temperate streams, rendering them impractical for such systems.

Suitable primary consumer taxa to use as baseline indicators in temperate streams have only been identified for Danish lowlands and for an eastern Canadian watershed (Anderson & Cabana, 2007; Kristensen et al., 2016). But few detailed comparative tests exist, and a consensus needs to be reached for baseline indicators across regions. Therefore, our objective was to compare macroinvertebrate taxa and groups of taxa (i.e., functional feeding group; FFGs) as baseline indicators that reflect d15N values at the base of food web in streams that span the Rocky Mountain—Great Plaines ecotone in western North America. These streams vary in land-use characteristics moving from upstream headwaters to downstream reaches, making them opportune systems to identify a widely useable and reliable baseline indicator. We used four criteria to identify an optimal invertebrate baseline (Kristensen et al., 2016). The baseline taxon must: (1) have a wide geographic distribution, (2) have low with-in site variation of d15N values, (3) have d15N values that are correlated with background spatial variability in d15N, and (4) give trophic position estimates of organisms higher in the food web that are free of environment influences of d15N. As urbanization and agricultural land-use practices can influence d15N values in aquatic ecosystems (Mayer et al., 2002; Anderson & Cabana, 2005), the identification of regionally-derived baseline indicator organisms should aid in quantifying fish trophic position in stable isotope studies in lotic systems.

@@

We sampled sixteen 2nd-5th order stream sites distributed along the Rocky Mountain—Great Plains ecotone. Sites were selected to represent an environmental gradient based on elevation and land use that resulted in a wide variation of d15N values at the base of local food webs. Sites spanned the longitudinal (i.e., upstream-downstream) gradient of the three primary tributaries to the North Platte River in Wyoming, USA (Figure 1). The Sweetwater, Medicine Bow and Laramie rivers are snowmelt dominated with peak flows occurring in the spring and base flows occurring in mid-summer. Their watersheds transition from forested land at high elevations to agriculturally dominated land with some urbanization at low elevations.

@@

Instead of using percent natural land cover as a variable to compare d15N values among potential baselines (Kristensen et al., 2016), we used a proxy variable for a site’s location along the longitudinal stream gradient developed by Maitland (2020). The proxy variable combines seven metrics of environmental conditions at a study site into an index of longitudinal stream position using the Axis 1 site score from a principal component analysis (PCA). Input variables to the PCA were elevation (m a.s.l.), stream slope (km/km), distance to North Platte River (km), Strahler stream order, mean August water temperature (℃), upstream drainage area (km2), and stream width (m). This longitudinal gradient index (PC1) explained 78% of site-level variation among the 16 study sites and described changes in environmental conditions from upstream, high-elevation, cold-water streams to downstream, low-elevation, warm-water streams. Upstream sites were more likely to be surrounded by natural land cover (e.g. forest, grassland, shrubland, and wetland land types; NLCD Database 2011), as evidenced by a strong negative correlation between PC1 and percent natural landcover at a site (Pearson's correlation, r = -0.69, t15 = -3.66, P < 0.001; Maitland, 2020). Maitland (2020) also found that PC1 (and thus the degree of natural landcover) was positively correlated with d15N values of basal resources and primary consumers (Linear Regression - Basal Resources: F1,373 = 121.4, P < 0.001; Primary Consumers: F¬¬1,14 = 4.7, P = 0.050), consistent with previous research (Cabana and Rasmussen 1996, Lake et al. 2001, Kristensen et al. 2016). Thus, we used PC1 as a measure of local variation in d15N values for all future analyses. 
Sample collection and preparation

Samples were collected during summer 2016. Basal resources (i.e., primary uptake compartments) were sampled once in June, July, and August to account for fine-scale temporal variation in d15N (Jardine et al., 2014).  Seston was collected by filtering three replicates of up to 10 L of water onto pre-combusted (550 °C, 4 hours) filters (Whatman GF/F, 47 mm O) in the field using a modified portable drill pump (Kelso & Baker, 2016). Biofilm was collected by scraping five cobble-sized rocks within the reach.  Macroinvertebrates were collected during the same sampling events during the summer of 2016. A D-framed kick net was used in a variety of habitats (i.e. riffle, pools, macrophytes, and river margins) to collect a representative sample of the macroinvertebrate assemblage.  Macroinvertebrates were kept in filtered stream water for 24 hours to allow for gut clearance, then frozen. Muscle tissue samples were obtained from fish collected during the August sampling event using a backpack electrofishing unit (Smith-Root, Vancouver, WA).  Brown Trout Salmo trutta (Linnaeus, 1758), Creek Chub Semotilus atromaculatus (Person, Year), White Sucker Catostomus commersonii (Lacépède, 1803), were used because these species were found at greater than 50% of the sites.  To minimize the number of fish euthanized, a muscle plug using a 5-mm biopsy punch was taken from fish greater than 300 mm in length. The remaining fish used for SIA were euthanized with a lethal dose of MS-222, placed on ice and then frozen upon returning to the lab. Those fish not selected for SIA were returned to the stream.

@@

When possible, we used individual macroinvertebrates for SIA but for small taxa we pooled individuals to ensure enough material for SIA. Muscle fillets were removed from the cranial dorsal portion of each fish. Samples were oven died (60 $^\circ$C, 48 hours), ground into a homogenous powder, and weighed to the nearest 0.001 mg. Animal samples (~1.0 mg) and basal resources (~2.0 mg) were weighed into 8 x 5-mm tin capsules and analyzed for d15N, d13C, %C, %N and C:N ratio at the University of Wyoming Stable Isotope Facility using a Delta Plus XP Continuous Flow Stable Isotope Ratio Mass Spectrometer (Thermo Finnigan, Bremen, Germany) coupled to a Costech Analytical 4010 elemental analyzer. Carbon (13C:12C) and nitrogen (15N:14N) isotope ratios (R) were estimated relative to their respective International Atomic Energy Agency standards (Pee Dee Belemnite limestone and atmospheric nitrogen). Working internal standards were run as controls throughout analysis to ensure accurate measurements. Isotopic values are reported in per mill delta notation \@ref(eq:ratio):

@@

where dN is either N15 or c13, and R is the ratio of the heavy to lighter isotope. Analytical error (i.e., 1 SD of lab standard) of sample runs was estimated at 0.07‰ (d13C) and 0.18‰ (d15N) for the invertebrates, and 0.04‰ (d13C) and 0.10‰ (d15N) for the fish.

@@

Suitability of various macroinvertebrate taxa as baselines were based on the four criteria proposed by Kristensen et al. (2016).  The baseline taxon must: (1) have a wide geographic distribution, (2) have low with-in site variation of d15N values, (3) have d15N values correlated with environmental influences of d15N, (4) give trophic positions estimates of other consumers free of environment influences of d15N.  Although all analyses were performed for all macroinvertebrate taxa and FFGs, selection occurred progressively from each criterion. For example, if a macroinvertebrate taxon met the first criterion it would stay in the list of potential baselines and be evaluated for the second criterion, but if it did not meet the first criterion then it would be removed from the list.

@@

We calculated the coefficient of variation (CV) of d15N values for all macroinvertebrate taxa and FFGs. A low coefficient of variation for d15N indicates that a baseline candidate has a non-omnivorous diet. To calculate CV, we divided the standard deviation of d15N signatures by the mean d15N signature of each baseline candidate at each site.  We then found the mean and 95% confidence intervals for each baseline candidate across sites.

@@

For all macroinvertebrate taxa and FFGs, we used least squares linear regression to assess relationships between the baseline’s d15N position at sites along the longitudinal stream gradient.  We wished to compare our baselines’ relationships to the longitudinal gradient with those of the basal resources. So, we did separate least squares linear regressions for two of the basal resources at our sites and compared R2 values between the baselines and the resources at our sites.  The R2 value of our basal resources represent the minimum level of variation that our alternate baselines should explain in order to be used as alternatives to the traditional method (e.g. all basal resources across a season).   Baselines with significant ($P \le 0.05$) relationships with the longitudinal gradient and that had R2 values equal to or greater than the basal resources were considered suitable baselines.

@@

where TPConsumer is the trophic position of each individual fish at a site, d15NConsumer is the d15N signature of each individual fish, d15NBaseline is the mean d15N signature of the baseline at each site, ∆15N is the discrimination factor for each trophic transfer, and TLBaseline is the expected trophic level of the baseline. Discrimination factor were estimated using the methods of Caut et al. (2009).

@@

We calculated the coefficient of variation (CV) for d15N at each site for the 12 sufficiently distributed baselines (8 taxa, 4 FFGs).  We found significant differences in the CVs for the taxa (ANOVA test; F11,26 = 2.96, P = 0.002; Figure 2). Simuliidae had the smallest CV across sites and Elmidae-adult had the largest (Figure 2).  Additionally, Simuliidae were 0.19 units lower than Elmidae-adult, 0.17 units lower than Chironimidae, and, although marginally significant, 0.15 units lower than Elmidae-larvae (Tukey HSD tests; P = 0.011 for Elmidae-adult; P = 0.032 for Chironomidae; P =  0.071 for Elmidae-adult; Figure 2).  Simuliidae were not statistically different from Dytiscidae, Baetidae, Heptaganeidae and Hydropyschidae. We found marginally significant differences among the FFG baselines (ANOVA test; F3,57 = 2.49, P = 0.070; Figure 2). Filterers were 0.18 units lower than Grazers (Tukey HSD tests; P = 0.047; Figure 2).  Filterers were not statistically different from Predators or Collectors.

### Criteria 3: Correlation with Environmental d15N

The two basal resources, biofilm and seston, had d15N signatures that were significantly (P ≤ 0.05) correlated with the longitudinal stream gradient (PC1) (Figure 3; Table 1). All eight of the well distributed taxa and all FFG’s had d15N signatures that were significantly correlated with PC1 (Figure 3). The amount of variation in d15N signatures captured by the longitudinal gradient was similar among the two basal resources, eight sufficiently distributed taxa and the four FFGs.

### Criteria 4: Trophic Position Estimates Independent of Environmental d15N

@@

Similar to FFG’s, correction of the TP using different taxa significantly reduced the effect of the longitudinal gradient on TP estimates (Figure 5; Figure 7; Supplemental Table 1).  The slope of Brown TP estimates and PC1 was lower for all taxa than when using uncorrected estimates (Figure 7). After correction, Brown Trout TP estimates were positively correlated with the longitudinal gradient when using Hydropyschidae, Baetidae, Simulidae, Dytiscidae, Elmidae-larvae, and Elmidae-adult.  Brown Trout TP estimates were not correlated with the longitudinal gradient when using Heptaganeidae and Chironomidae as baselines. Brown Trout trophic position estimates differed by baseline taxa (ANOVA test: F7,2787 = 68.76; P < 0.001).

Brown Trout TP estimates given by Elmidae-adult and Elmidae-larvae were higher than the other taxa (Figure 6; Supplemental Table 2). Hydropyschidae and Heptaganeidae gave lower Brown Trout TP estimates than the other except Chironomidae and Baetidae for which they did not differ (Figure 6; Supplemental Table 2).  Dytiscidae and Simulidae gave intermediate Brown Trout TP estimates (Figure 6; Supplemental Table 2).

@@

Figure 2. Each baseline’s mean Coefficient of Variation (CV) of d15N across sites with 90% confidence intervals. A) Taxon baselines B) FFG baselines.  All baselines met Criteria 1 of being found in at least 75% of study sites.

Figure 3. Relationships between the d15N values of A) basal resources, B) FFGs, and C) taxa in relation to position along the longitudinal gradient represented by PC1.  Each point is the mean d15N signature of the baseline candidate or basal resource at that site. Upstream sites are to the left. The black line is the fitted regression line and the shaded grey represents the 95% confidence interval.  See Supp. Table 1 for full statistical results.
